\pkg{digiRhythm}: an \proglang{R} package for rhythmic behavior assessment in livestock

### 2023-05-17 jstatsoft/article.Rnw
changes marianne and marie

The real-time monitoring of animal behaviour is becoming increasingly important in agriculture. It can provide new opportunities to enhance the efficiency and sustainability of livestock production and is used to enhance animal welfare (Wemelsfelder et al., 2001). Animal behaviour can be altered by stress, pain or disease, but monitoring these changes in welfare is still a challenge, since some behavioural definitions are unclear and not specifically linked to a unique adverse event (Theurer et al., 2013; Levitis et al., 2009). Relevant information about animal welfare can be obtained by studying individual differences in the capacity for behavioural adaptation (Cooper et al., 2005). Indeed, during the animals life, significant changes can occur due to internal and external stimuli; these alterations arouse adaptive behavioural patterns that can be monitored by analysing the rhythmicity of behaviours (Berger et al., 2003).
Due to the increasing development of ICT (Information and Communication Technology) based on small, non-invasive sensors, large amounts of animal behaviour data are regularly recorded, and can be used to calculate parameters describing biological rhythms (Theurer et al., 2013; Scheibe and Gromann, 2006; Berger et al. 2003). Scheibe et al. (1999) developed a parameter able to quantify and investigate the rhythmicity of behavioural patterns, the Degree of Functional Coupling (DFC), which is based on telemetric measurements of activity and/or feeding. The DFC provides an estimate of synchronization between an individual and its environment. It is calculated to each 24 h period within rolling seven day periods and predicts values between $0\%$
and 100 $\%$ (Scheibe et al., 1999, Berger et al., 2003). An organism with a high DFC is highly synchronized with its environment, while lower DFC's indicate a lower synchronization. Breakdowns in the DFC response can be caused by stressors (e.g. transport, disease, etc.) that affect the time structure of behaviour (Berger et al., 2003; Scheibe et al., 1999). In fact, animals that live in stress-free conditions show activities all tuned to one zeitgeber (i. e. 24 hour periodicity), while their activity rhythm will deviate from the zeitgeber frequency if they are subjected to stressful situations (Scheibe et al., 1999; Berger et al., 2003).
Previous studies demonstrated that calves change their behaviour as illness develops (reduced feeding, less social behaviour and increased rest) as an adaptive response to cope with the disease (Owen-Ashley et al., 2006, Dantzer and Kelley, 2007; Borderas et al. 2008). A better understanding of changes in behaviour due to illness or stress of handling is valuable to improve animal welfare and productivity as well as handler safety (Grandin et al., 1989; Kilgore et al., 2005; Borderas et al., 2008; Robért et al., 2011). Bio-rhythmic analysis has the potential to provide the basis for the early detection of psychological and physiological disorders (Berger et al., 1999).

With the increasing availability of sensor technologies to collect behavioral data continuously, further studies applying the DFC were performed \citep{sarout2018assessment, fuchs2022detecting}. These studies were able to confirm that the rhythm of animals is linked to their state of physical wellbeing. However, the application of the DFC was always limited to those with extensive knowledge of statistical programming. An open source package specifically for this type of computation opens up the opportunities for livestock research to easily apply these algorithms. Therefore, we developed the \proglang{R} package \pkg{digiRhythm} \citep{dgm}.

@@

The function \fct{sliding\_DI} also computes the diurnality index as described above. The difference is that no single definition of day and night was used for the complete calculation period. It is possible to include a data frame with different definitions of the start/end of day/night (namely, the variable $day\_start$, $day\_end$, and $night\_start$ and $night\_end$). The data frame should include five columns, one date column, and four columns of timestamps for start/end of day/night. All of these columns should be in the POSIXct format.

This method, which includes different definitions, could be helpful in various experiments. For example, in Europe, it could be used to reduce the bias emerging by converting summertime and wintertime to a time zone, such as GMT. An additional use could be including sunrise and sunset by using the same definition for $day\_start$, $day\_end$, and $night\_start$ and $night\_end$.

@@

\section{Discussion} \label{sec:summary}
The \pkg{digiRhythm} package can make a significant contribution to the assessment of rhythmicity in animals, as it makes the analysis of animal rhythmicity more accessible to researchers. The tool is easy to use, especially because it is delivered as an \proglang{R} package. The theoretical details and intuition behind the DFC is explained in section \ref{sec:dfc}. The usage of the package is explained in the section \ref{sec:usage}. Further usage examples can be found in the library's vignettes.

The package enables the calculation of rhythmicity in, but not limited to, animals. To date, this is purely a computation of activity patterns that have been correlated with physiological welfare factors in previous studies. This indicator can be calculated from any type of regularly measured time series data. However, we have no evidence how these activity patterns are regulated by routine management practices, such as feeding or milking. Such routines could indirectly affect rhythmic activity patterns and blur physiological rhythmicity patterns. To date, the package provides the option to exclude known management time spans from the analysis. However, in order to increase the understanding of rhythmicity as a welfare indicator future studies should investigate how these management practices affect rhythmicity calculations and evaluate if this is a reasonable welfare indicator for animals depending on external management.

The required frequency of data collection is of utmost importance in animal-borne sensor technologies. Sensors are required to be small, so that they don't disturb the animal, and further expected to run without any sensor based maintenance, such as data reading or charging of the device. This in turn means that data collection frequency should be low. In the current study an extremely low data collection frequency is suggested (15 minutes) future studies should evaluate via downsampling, how low collection frequency can be set to create reliable rhythmicity calculations. Further it would be interesting to evaluate if activity data collected from different sources, such as access of feeding stations or watering troughs, compute similar rhythmicity outputs in the same organism.

Include 7 days aspect.

>>>>

Internal and external stimuli arouse adaptive behavioural patterns that can be monitored by analysing the rhythmicity of behaviours (Berger et al., 2003). Due to the increasing development of ICT (Information and Communication Technology) based on small, non-invasive sensors, large amounts of behavioural data are regularly recorded, and can be used to calculate parameters describing biological rhythms (Theurer et al., 2013; Scheibe and Gromann, 2006; Berger et al. 2003). \cite{sinz1976systemanalyse} developed a parameter able to quantify and investigate the rhythmicity of behavioural patterns, the Degree of Functional Coupling (DFC), which is based on telemetric measurements of activity and/or feeding of animals. The DFC provides an estimate of synchronization between an individual and its environment. It is calculated to each 24 h period within a timespan of a few days and predicts values between $0\%$ and 100 $\%$ \citep{scheibe1999comparative}, Berger et al., 2003). XXX suggested to use rolling seven day periods for this computation. An organism with a high DFC is highly synchronized with its environment, while lower DFC's indicate a lower synchronization. Breakdowns in the DFC response can be caused by stressors (e.g. handling, disease, etc.) that affect the time structure of behaviour (Berger et al., 2003; Scheibe et al., 1999). In fact, animals that live in stress-free conditions show activities all tuned to one zeitgeber (i. e. 24 hour periodicity), while their activity rhythm will deviate from the zeitgeber frequency if they are subjected to stressful situations (Scheibe et al., 1999; Berger et al., 2003).
A better understanding of changes in behaviour due to illness or distress is valuable to get further inisight in the organisms affective state % checke reference
(Grandin et al., 1989; Kilgore et al., 2005; Borderas et al., 2008; Robért et al., 2011). Bio-rhythmic analysis has the potential to provide the basis for the early detection of psychological and physiological disorders (Berger et al., 1999). With the increasing availability of sensor technologies to collect behavioral data continuously, further studies applying the DFC were performed \citep{sarout2018assessment, fuchs2022detecting}. These studies applied the DFC method in managed animal husbandry systems. However, the application of the DFC was always limited to those with extensive knowledge of statistical programming. An open source package specifically for this type of computation opens up the opportunities for livestock research to easily apply these algorithms. Therefore, we developed the \proglang{R} package \pkg{digiRhythm} \citep{dgm}, which was originally developed for livestock research, but has the potential to be applied in various domains.

@@

The function \fct{sliding\_DI} computes the diurnality index as described above. However, instead of using a single definition of day and night for the complete computation period, a data frame defining different definitions of the start/end of day/night is included. The data frame must include five columns, one date column, and four columns of timestamps for start/end of day/night. All of these columns must be formatted as POSIXct.

This function, which includes different definitions, is useful in various settings, as it allows to improve the computation accuracy. For example, in Europe, it could be used to reduce the bias emerging by converting summertime and wintertime to a time zone, such as GMT, or by defining sunrise and sunset.

@@

\section{Discussion} \label{sec:summary}
The \pkg{digiRhythm} package can make a significant contribution to the assessment of rhythmicity, as it makes the analysis of rhythmicity more accessible to researchers. The tool is easy to use, especially because it is delivered as an \proglang{R} package. The theoretical details and intuition behind the DFC is explained in section \ref{sec:dfc}. The usage of the package is explained in the section \ref{sec:usage}. Further examples can be found in the library's vignettes.

The package enables the calculation of rhythmicity from activity data. To date, this is purely a computation of activity patterns that have been correlated with physiological factors in previous studies. This indicator can be calculated from any type of regularly measured time series data, if they are brought in a suitable format, which can be assessed with the  \fct{is_dgm_friendly} function. However, we have no evidence if these activity patterns are intrinsically (such as going to sleep) or extrinsically (such as waking up to an alarm) motivated. Such extrinsic factors could affect rhythmic activity patterns and blur physiological rhythmicity patterns. To date, the package provides the option to exclude extrinsically structured time spans from the analysis. However, in order to increase the understanding of rhythmicity  future studies should investigate how these extrinsic factors affect rhythmicity calculations.

The required frequency of data collection is of utmost importance in sensor technologies. Sensors are required to be small and further expected to run without any sensor based maintenance, such as data reading or charging of the device. This in turn means that data collection frequency should be low. In the current study an extremely low data collection frequency is suggested (15 minutes) future studies should evaluate how low collection frequency can be set to create reliable rhythmicity calculations. Further it would be interesting to evaluate if activity data collected from different sources, such as  access of feeding stations or coffee maschines, compute similar rhythmicity outputs in the same organism.

Seven days of continuous data collection are required to compute one DFC value. This implies that very long data collection intervals are required to detect rhythmic patterns. Reducing the number of days required to compute one DFC value would make this rhythmicity computation much more accessible in practice an research. The impact of reducing the number of days required to compute the DFC should be investigated in future research.
